```{r, warning = FALSE}
library(ggplot2)
library(caret)
```

## Summarize data
```{r, warning = FALSE}
# Load the dataset 
dataset <- read.csv("Boulder Data.csv", header = TRUE)

# Load the location data
locationset <- read.csv("Boulder Location.csv", header = TRUE)

# Peak into the data
head(dataset)

# Convert from character string into a time/date object
dataset$Checkout.Date <- strptime(dataset$Checkout.Date, "%m/%d/%Y")
dataset$Checkout.Date <- format(dataset$Checkout.Date, "%m/%d/%Y")

# Convert from character string into a time/date object
dataset$Checkout.Time <- strptime(dataset$Checkout.Time, "%I:%M:%S %p")
dataset$Checkout.Time <- format(dataset$Checkout.Time, "%H:%M:%S")

# Get the summary of the data
summary(dataset)

# View the subset of the data
subset <- subset(dataset, dataset$Rider.Home.System == "Denver B-cycle")
```

From the summary it is clear that there are some errors in the "Rider.Home.System" column.
This is supposed to be data from Boulder, so there are some errors in terms of the home system.
Subsetting the data and then glancing at the Checkout station and Return station, it is clear
that the stations are correct and the home system is wrong. Not a big deal though because this 
variable/column data is probably not that important

## Visualize data
```{r, warning = FALSE}
# Bar plot of the post popular pass type
riders <- subset(dataset, 
                 Rider.or.Operator.Number%in%levels(dataset$Rider.or.Operator.Number)
                 [table(dataset$Rider.or.Operator.Number) >= 300])

ggplot(riders, aes(Rider.or.Operator.Number)) + geom_bar() + 
  facet_wrap(~Entry.Pass.Type, ncol = 1) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Bar plot of the post popular pass type
ggplot(dataset, aes(Entry.Pass.Type)) + geom_bar()
```

